---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# panelTVP

Hello and welcome to the package panelTVP! 

This package provides functions for fitting Bayesian regression models for panel data with time-varying parameters. The functions in this package were created by Roman Pfeiler and Helga Wagner. Maintainer of the package is Roman Pfeiler (roman.pfeiler@jku.at).

## Installation

You can install the development version of panelTVP from Roman Pfeiler's personal GitHub page [GitHub](https://github.com/pfeilR/) 

``` r
install_github("pfeilR/panelTVP")
```

## Introduction to the package

The main goal of this package is to fit flexible Bayesian regression models for panel data in the time-varying parameter framework. The model is flexible in the sense that for each covariate and the subject-specific random effect, time-varying effects are estimated. For the random effects, a factor model specification is used, where time-varying factor loadings are multiplied with subject-specific factor scores. This strategy allows estimation of time-varying random effects. To avoid overfitting, hierarchical shrinkage priors are used to identify whether an effect is time-varying, time-invariant or zero. The following models are implemented:

* Gaussian model for normally distributed responses
* Logit / Probit model for binary responses
* Negative Binomial model for overdispersed count data
* Zero-Inflated Negative Binomial model for zero-inflated and overdispersed count data

## Small case-study: Gaussian response data

In the following, we show how to fit a Bayesian time-varying parameter model for panel data, where the conditional distribution of the response variable is assumed to be Gaussian. For simplicity, the statistical analysis is based on simulated data. 

Data can be simulated by using the function sim_panelTVP() from the panelTVP package as follows:

```{r}
devtools::load_all()
d <- sim_panelTVP(n = 1000,
                  Tmax = 4,
                  model = "Gaussian",
                  beta = c(4,1,0),
                  theta = c(1,0.5,0),
                  lambda = 1,
                  psi = 0.2,
                  sigma2 = 3)
```

Object **d** contains the generated dataset that includes the covariates and columns with time and subject indices, respectively:

```{r}
head(d$observed, 10)
```

The time varying regression effects and factor loadings can be accessed as follows:

```{r}
d$beta
d$lambda
```

It should be noted that the first covariate effect corresponds to the global intercept.

For fitting a Bayesian time-varying parameter model to the panel data, the main function panelTVP() is used. The function contains a lot of arguments, which are explained in greater depth in the corresponding help page. Moreover, only a very small number of draws in MCMC estimation are used here as this is only an illustrative example. In applications, you should opt for a longer Markov Chain.

```{r}
res <- panelTVP(y ~ W1 + W2,
                data = d$observed,
                id = d$observed$id,
                t = d$observed$t,
                model = "Gaussian",
                mcmc.opt = list(chain.length = 500, burnin = 100, thin = 2, asis = TRUE))
```

A summary table of the estimated coefficients can easily be created by calling the corresponding S3-function:

```{r}
summary(res)
```

Alternatively, results can also be ordered by time point:

```{r}
summary(res, by = "timepoint")
```

Covariate effect plots - that include the estimated factor loadings as well - are created as follows:

```{r}
plot(res, nplots = 2)
```

As we can see, the insignificant effect of covariate **W2** is effectively shrunken towards zero, while the other effects vary over time - as we would expect based on how the data were generated.

## The end?

Note that this was just a very simple and quick example for the sake of creating a first overview of the package. For a more thorough introduction to the package visit the packages' Vignette under **INSERT LINK TO VIGNETTE**.

